---
output:
  html_document:
    includes:
      in_header: hero-image.html
    css: styles.css
---

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
library(tidyverse)
library(fs)
library(glue)
library(lubridate)
```

<br>

# About
Hi there, I’m Laurence and welcome to my blog. I made this blog to showcase projects that I enjoy, to highlight the beauty of the R programming language from my perspective and also as a medium for myself and others to learn. This blog was created entirely in R Markdown and the source code can be found [here](https://github.com/ljwoodley/ljwoodley.github.io){target="_blank"}.

I graduated from the University of Florida with an undergraduate degree in Statistics and since then I have had the pleasure of working with a wide variety of data from political, health and criminal intelligence domains. Currently, I work at the University of Florida Clinical and Translational Science Informatics department where I spend my days working on projects centered around analytics engineering. The majority of my work is done with R, MySQL and REDCap but I also use Docker and Python when necessary.

When I’m not working with R I enjoy reading books, especially statistics books like Naked Statistics, exploring data science blogs and traveling.

```{r  message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
all_posts <- dir_tree("posts", glob = "*.html|*.txt") %>%
  tibble() %>%
  select("file_name" =  1) %>% 
  separate(
    file_name,
    c("main_folder", "post_date", "file"),
    sep = "/",
    remove = FALSE
  ) %>% 
  filter(!is.na(file)) %>% 
  arrange(desc(post_date)) %>% 
  group_by(file_name) %>%
  mutate(
    post_description = case_when(
      str_detect(file_name, "description.txt") ~ read_file(file_name),
      TRUE ~ NA_character_
    ),
     post_title = case_when(
      str_detect(file_name, "post_title.txt") ~ read_file(file_name),
      TRUE ~ NA_character_),
    post_date = format(ymd(post_date), format = '%b %d, %Y' )
  ) %>%
  group_by(post_date) %>% 
  fill(c(post_description, post_title), .direction = "downup") %>%
  filter(str_detect(file_name, ".html"))
```

<br>

# Posts
```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, results='asis'}
glue_data(all_posts,
    "<h3><a class='links' href='{file_name}'>{post_title}</a></h3>",
    "<h5>{post_date}</h5>",
    "<p>{post_description}</p>",
    "<a class='btn btn-outline-primary btn-sm' href='{file_name}'>Read More</a>",
    "<hr>"
  )
```


