---
title: 
output:
  html_document:
    includes:
      in_header: ../navbar.html
    css: ../styles.css
    highlight: pygments
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
library(bigrquery)
library(DBI)
library(tidyverse)
library(kableExtra)
library(knitr)
library(reticulate)
```


```{r include=FALSE}
con <- dbConnect(
  bigrquery::bigquery(),
  project = "ljwblog",
  dataset = "challenge_2"
)
```

# Query Challenge #2
 
The __option_transactions__ table shows options trades executed on the thinkorswim trading platform. Query the table to calculate the profit/loss made on each trade.

```{sql connection=con, include=FALSE, output.var='option_transactions'}
select * from option_transactions
```

```{r echo=FALSE}
kable(option_transactions, caption = "option_transactions") %>% 
  kable_styling(full_width = FALSE)
```

The anatomy of an options quote can be found
[here](https://www.ally.com/do-it-right/investing/trading-options-for-beginners/){target="_blank"}
while this
[video](https://www.youtube.com/watch?v=ciKRHrdFe1U){target="_blank"}
provides additional information on how to read an options quote.

Before getting into the query let's disect `TOSWeb BOT +1 RBLX 100 (Weeklys) 13 AUG 21 85 CALL @.82` to ensure that it's understandable:

1.   `TOSWeb` - The thinkorsiwm platform. This can be ignored
2.  `BOT` - The type of transaction. Can be either "BOT" (contract was bought) or "SOLD" (contract was sold)
3.  `+1` - The number of contracts bought or sold
4.  `RBLX` - The underlying stock symbol
5.  `100 (Weeklys)`  - Can be ignored as it does not provide any needed information
6.  `13 AUG 21` - The date that the contract expires
7.  `85` - The strike price of the stock
8.  `CALL` - The type of option. Can be either "CALL" or "PUT"
9.   `@.82` - The per contract cost of the option

The first two rows in the __option_transactions__ table shows that 2 RBLX calls 
with a strike pirce of \$85 and an expiration of 08/13 were bought
for a total price of \$1.82. The third row shows that both contracts
were sold for a price of \$2.10 per contract. Thus, the total profit made on this trade was \$2.38 

#  {.tabset}

## SQL

```{sql connection=con, output.var='profit_loss'}
with clean_options as (
  select
    stock, 
    transaction_type, 
    contract,
    sum(num_contracts * price) as total_price,
  from (
    select
        option_quote,
        trim(regexp_extract(option_quote, r"(.*? ){2}")) as transaction_type,
        cast(regexp_extract(option_quote, r"\d+") as numeric) as num_contracts,
        regexp_extract(option_quote, r"(.*? ){4}") as stock,
        regexp_extract(option_quote, r"\(Weeklys\)?(.*)@") as contract,
        cast(regexp_extract(option_quote, r"\@(.*)$") as numeric) as price
    from option_transactions)
  group by 
    stock, 
    transaction_type, 
    contract
),

buy_to_open as (
    select
        stock,
        contract,
        total_price as buy_price
    from clean_options
    where transaction_type = 'BOT'
),

sell_to_close as (
    select
        stock,
        contract,
        total_price as sell_price
    from clean_options
    where transaction_type = 'SOLD'
)

select
    buy_to_open.stock,
    buy_to_open.contract,
    buy_to_open.buy_price,
    sell_to_close.sell_price,
    (sell_to_close.sell_price - buy_to_open.buy_price) as profit_loss
from
  buy_to_open
  inner join sell_to_close on 
    buy_to_open.stock = sell_to_close.stock and
    buy_to_open.contract = sell_to_close.contract
order by profit_loss desc
```

```{r echo=FALSE}
kable(profit_loss) %>% 
  kable_styling(full_width = FALSE)
```

```{r include=FALSE}
dbDisconnect(con)
```

## R

```{r}
clean_options <- option_transactions %>%
  mutate(
    transaction_type = word(option_quote, 2),
    num_contracts = abs(parse_number((word(option_quote, 3)))),
    stock = word(option_quote, 4),
    contract = word(option_quote, 7, 11),
    price = parse_number(word(option_quote, -1)),
    total_price = num_contracts * price
  ) %>% 
  group_by(stock, transaction_type, contract) %>% 
  summarise(total_price = sum(total_price)) %>% 
  ungroup()

buy_to_open = clean_options %>% 
  filter(transaction_type == 'BOT') %>% 
  select(stock, contract, buy_price = total_price)

sell_to_close = clean_options %>% 
  filter(transaction_type == 'SOLD') %>% 
  select(stock, contract, sell_price = total_price)

profit_loss <- buy_to_open %>% 
  inner_join(sell_to_close, by = c("stock", "contract")) %>% 
  mutate(profit_loss = sell_price - buy_price) %>% 
  arrange(desc(profit_loss))

```

```{r echo=FALSE}
kable(profit_loss) %>% 
  kable_styling(full_width = FALSE)
```

## Python

```{python}
option_transactions = r.option_transactions

clean_options = (
  option_transactions
  .assign(
    option_quote = lambda x: x["option_quote"].str.split(" "),
    transaction_type = lambda x: x["option_quote"].str[1],
    num_contracts = lambda x: x["option_quote"].str[2].apply(int).abs(),
    stock = lambda x: x["option_quote"].str[3],
    contract = lambda x: x["option_quote"].str[6:10].str.join(" "),
    price = lambda x: x["option_quote"].str[-1].str.replace("@", "").apply(float),
    total_price = lambda x: x["num_contracts"] * x["price"]
    )
  .groupby(['stock','transaction_type','contract'])['total_price']
  .sum()
  .reset_index()
  )

buy_to_open = (
  clean_options[clean_options['transaction_type'] == 'BOT']
  [['stock', 'contract', 'total_price']]
  .rename({"total_price": "buy_price"}, axis='columns')
  )
  
sell_to_close = (
  clean_options[clean_options['transaction_type'] == 'SOLD']
  [['stock', 'contract', 'total_price']]
  .rename({"total_price": "sell_price"}, axis = 'columns')
  )

profit_loss = (
  buy_to_open
  .merge(sell_to_close, how='inner', left_on=['stock', 'contract'], right_on=['stock', 'contract'])
  .assign(profit_loss = lambda x: x["sell_price"] - x["buy_price"])
  .sort_values(by='profit_loss', ascending=False)
  )
```

```{r echo=FALSE}
kable(py$profit_loss) %>% 
  kable_styling(full_width = FALSE)
```
